Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-09 analysis/paper/paper.Rmd
ACKNOWLEDGEMENTS UPDATE PER MFC

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
2014) have used estimated haul-out probabilities to correct for the proportion of animals
that are in the water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by day-of-year and
time-of-day, previous correction factors have not accounted for variability between years,
different weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled out on ice
based on these factors. We thus recommend that future abundance analyses employ correction
factors that account for these factors. For instance, it is relatively straightforward to
obtain weather reanalysis products for times and locations that are surveyed and to
construct a relevant correction factor based on predictions of GMPMs. Further, one can
apply stable stage distribution estimates from matrix population models to adjust
predictions to the "average" prediction of the population. The latter suggestion is
particularly appropriate when the age structure of the tagged sample does not reflect the
likely age structure of the population. For instance, population weighted estimates can
differ substantially from those based on a naïve analysis of the tagged sample, as was the
case with analysis of spotted seal data (e.g. Fig X). The most challenging element in
developing availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors, particularly
because research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One possible suggestion
is to estimate a "shift" parameter within models for aerial survey counts that allow the
peak of haul-out distributions to be adjusted earlier or later in the year based on the
frequency of counts observed over time. Regardless, researchers should anticipate there
being some un-modeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.

@@

*Josh M. London:* led field work, conceptualization, methodology, formal analysis, code development, writing of original draft, visualization, review, revisions, and corresponding author

*Paul B. Conn:* conceptualization, methodology, formal analysis, code development, writing of original draft, review, and revisions

*Stacie K. Hardy:* field work, methodology, data curation, validation, review, and revisions

*Erin L. Richmond:* data curation, conceptualization, methodology, review, and revisions

*Jay M. Ver Hoef:* field work, conceptualization, methodology, code development, review, and revisions

*Irina S. Trukhanova:* conceptualization, formal analysis, code development, review, and revisions

*Michael F. Cameron:* led field work, project administration, methodology, review, and revisions

*Peter L. Boveng:* led field work, supervision, project administration, conceptualization, methodology, review, and revisions

@@

We recognize that the species and ecosystems studied in this paper are within the
traditional range of the Inpuiaq and Yupik people. And, we would like to acknowledge and
thank them as the original stewards of these ecosystems. The deployment of bio-logging
devices used in this study were often done in collaboration with Alaska Native
communities. We would like to especially acknowledge the communities of Kotzebue,
Utqiagvik, and Ulguniq.

We are grateful for the assistance in catching and sampling seals by Michelle Barbieri,
Gavin Brady, Cynthia Christman, Shawn Dahle, Stacy DiRocco, Deb Fauquier, Shannon
Fitzgerald, Tracey Goldstein, Jeff Harris, Markus Horning, John Goodwin, John Jansen,
Shawn Johnson, Charles Littnan, Erin Moreland, Lorrie Rea, Dan Savetilik, and Dave
Withrow. We also appreciate the commitment to science and safety by all officers and crew
of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.

>>>>

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
2014) have used estimated haul-out probabilities to correct for the proportion of animals
that are in the water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by day-of-year and
time-of-day, previous correction factors have not accounted for variability between years,
different weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled out on ice
based on these factors. We thus recommend that future abundance analyses employ correction
factors that account for these factors. For instance, it is relatively straightforward to
obtain weather reanalysis products for times and locations that are surveyed and to
construct a relevant correction factor based on predictions of GMPMs. Further, one can
apply stable stage distribution estimates from matrix population models to adjust
predictions to the "average" prediction of the population. The latter suggestion is
particularly appropriate when the age structure of the tagged sample does not reflect the
likely age structure of the population. For instance, population weighted estimates can
differ substantially from those based on a naïve analysis of the tagged sample, as was the
case with analysis of spotted seal data (e.g. Fig X). The most challenging element in
developing availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors, particularly
because research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One possible suggestion
is to estimate a "shift" parameter within models for aerial survey counts that allow the
peak of haul-out distributions to be adjusted earlier or later in the year based on the
frequency of counts observed over time. Regardless, researchers should anticipate there
being some unmodeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.

@@

**Josh M. London:** led field work, conceptualization, methodology, formal analysis, code development, writing of original draft, visualization, review, revisions, and corresponding author

**Paul B. Conn:** conceptualization, methodology, formal analysis, code development, writing of original draft, review, and revisions

**Stacie K. Hardy:** field work, methodology, data curation, validation, review, and revisions

**Erin L. Richmond:** data curation, conceptualization, methodology, review, and revisions

**Jay M. Ver Hoef:** field work, conceptualization, methodology, code development, review, and revisions

**Irina S. Trukhanova:** conceptualization, formal analysis, code development, review, and revisions

**Michael F. Cameron:** led field work, project administration, methodology, review, and revisions

**Peter L. Boveng:** led field work, supervision, project administration, conceptualization, methodology, review, and revisions

@@

We recognize that the species and ecosystems studied in this paper are within the
traditional range of the Inpuiaq and Yupik people. And, we would like to acknowledge and
thank them as the original stewards of these ecosystems. The deployment of bio-logging
devices used in this study were often done in collaboration with Alaska Native
communities. We would like to especially acknowledge the communities of Kotzebue,
Utqiagvik, and Ulguniq and the following individuals: James Adams, Jeff Barger, David
Barr, Wendell Booth, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John
Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr., Virgil
Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik from Kotzebue, Alaska; Billy
Adams, Mary-Ellen Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard
Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and Joe Skin
from the North Slope Borough, Alaska;

We are grateful for the assistance in catching and sampling seals by James Bailey,
Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov, Cynthia Christman, Shawn
Dahle, Rob Delong, Stacy DiRocco, Deb Fauquier, Shannon Fitzgerald, Kathy Frost, Scott
Gende, Tracey Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson, Charles
Littnan, Lloyd Lowry, Erin Moreland, Lorrie Rea, Gay Sheffield, Brent Stewart, and Dave
Withrow. We also appreciate the commitment to science and safety by all officers and crew
of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.
